# Practical Machine Learning - Final Project

## Overview

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set. We train 4 models: Decision Tree, Random Forest, Gradient Boosted Trees, Support Vector Machine using k-folds cross validation on the training set. We then predict using a validation set randomly selected from the training csv data to obtain the accuracy and out of sample error rate. Based on those numbers, we decide on the best model, and use it to predict 20 cases using the test csv set.

## Loading Data and Libraries

```{r}
library(lattice)
library(ggplot2)
library(caret)
library(kernlab)
library(rattle)
library(corrplot)
library(e1071)
set.seed(1234)
```

```{r}
traincsv <- read.csv("./pml-training.csv")
testcsv <- read.csv("./pml-testing.csv")
```

```{r}
dim(traincsv)
```

```{r}
dim(testcsv)
```

## Cleaning the Data

Removing unnecessary variables. Starting with N/A variables.
```{r}
traincsv <- traincsv[,colMeans(is.na(traincsv)) < .9] #removing mostly na columns
traincsv <- traincsv[,-c(1:7)] #removing metadata which is irrelevant to the outcome
```

Removing near zero variance variables.
```{r}
nvz <- nearZeroVar(traincsv)
traincsv <- traincsv[,-nvz]
dim(traincsv)
```

Now that we have finished removing the unnecessary variables, we can now split the training set into a validation and sub training set. 
```{r}
inTrain <- createDataPartition(y=traincsv$classe, p=0.7, list=F)
train <- traincsv[inTrain,]
valid <- traincsv[-inTrain,]
```

## Creating and Testing the Models

Here we will test a few popular models including: Decision Trees, Random Forest, Gradient Boosted Trees, and SVM. 

Set up control for training to use 3-fold cross validation.
```{r}
control <- trainControl(method="cv", number=3, verboseIter=F)
```

### Decision Tree
```{r}
mod_trees <- train(classe~., data=train, method="rpart", trControl = control)
fancyRpartPlot(mod_trees$finalModel)
```

Prediction 
```{r}
pred_trees <- predict(mod_trees, valid)
cmtrees <- confusionMatrix(pred_trees, factor(valid$classe))
cmtrees
```

### Random Forest 
```{r}
mod_rf <- train(classe~., data=train, method="rf", trControl = control, tuneLength = 5)

pred_rf <- predict(mod_rf, valid)
cmrf <- confusionMatrix(pred_rf, factor(valid$classe))
cmrf
```

### Gradient Boosted Trees
```{r}
mod_gbm <- train(classe~., data=train, method="gbm", trControl = control, tuneLength = 5, verbose = F)

pred_gbm <- predict(mod_gbm, valid)
cmgbm <- confusionMatrix(pred_gbm, factor(valid$classe))
cmgbm
```

### Support Vector Machine
```{r}
mod_svm <- train(classe~., data=train, method="svmLinear", trControl = control, tuneLength = 5, verbose = F)

pred_svm <- predict(mod_svm, valid)
cmsvm <- confusionMatrix(pred_svm, factor(valid$classe))
cmsvm
```


The best model is the Random Forest model, with 0.9958 accuracy and 0.0042 out of sample error rate. We find that to be a sufficient enough model to use for our test sets.


## Predictions on Test Set

Running our test set to predict the classe (5 levels) outcome for 20 cases with the Random Forest model.

```{r}
pred <- predict(mod_rf, testcsv)
print(pred)
```

## Appendix
correlation matrix of variables in training set
```{r}
corrPlot <- cor(train[, -length(names(train))])
corrplot(corrPlot, method="color")
```

Plotting the models
```{r}
plot(mod_trees)
```

```{r}
plot(mod_rf)
```

```{r}
plot(mod_gbm)
```


